---
title: "R Notebook"
output: html_notebook
---

Install required packages

```{r}
install.packages(c("tidyverse", "stats", "margins"))
```

Load required packages

```{r}
library(tidyverse)  # For data manipulation and plotting
library(stats)      # For glm() - logit and probit models
```

Prepare the mtcars dataset

```{r}
data(mtcars)
mtcars <- mtcars %>% 
  mutate(am_binary = as.factor(am))  # Binary outcome: 0 = auto, 1 = manual
```

## 1. Logit Model (Logistic Regression)

What Are Logit and Probit Models?

Logit (Logistic Regression): Models the log-odds of an event (e.g., am = 1) using a logistic function. The probability curves into an S-shape, staying between 0 and 1.

```{r}
logit_model <- glm(am_binary ~ hp + wt, 
                   data = mtcars, 
                   family = binomial(link = "logit"))
summary(logit_model)
```

## 2. Probit Model

Probit: Models the same probability but assumes a normal distribution (cumulative distribution function, CDF) for the underlying latent variable. Also S-shaped, but slightly different curvature.

```{r}
probit_model <- glm(am_binary ~ hp + wt, 
                    data = mtcars, 
                    family = binomial(link = "probit"))
summary(probit_model)
```

**Key Difference:**

-   **Logit**: Uses the logistic distribution (fatter tails), coefficients relate to log-odds.

-   **Probit**: Uses the normal distribution, coefficients are in z-score units (harder to interpret directly).

## 3. Predicted Probabilities

```{r}
mtcars$logit_prob <- predict(logit_model, type = "response")  # Probabilities
mtcars$probit_prob <- predict(probit_model, type = "response")
```

## 4. Plot Logit vs. Probit Predictions

```{r}
ggplot(mtcars, aes(x = hp)) +
  geom_point(aes(y = am, color = "Actual"), alpha = 0.5) +
  geom_line(aes(y = logit_prob, color = "Logit"), size = 1) +
  geom_line(aes(y = probit_prob, color = "Probit"), size = 1, linetype = "dashed") +
  labs(title = "Logit vs. Probit: Probability of Manual Transmission",
       x = "Horsepower (hp)", y = "Probability") +
  scale_color_manual(values = c("Actual" = "black", "Logit" = "blue", "Probit" = "red")) +
  theme_minimal()
```

## 5. Marginal Effects (for Logit) - approximate change in probability

```{r}
library(margins)  # For marginal effects
logit_margins <- margins(logit_model)
summary(logit_margins)
```
