---
title: "R Notebook"
output: html_notebook
---

## Part 1: Setting up Data

Set a random seed for reproducibility.

```{r}
set.seed(123)

install.packages(c("AER", "plm", "tidyverse"))
```

Install required packages for Panel Data Analysis.

```{r}
install.packages(c("AER", "plm", "tidyverse"))
```

Load the required packages you've previously installed.

```{r}
library("AER")        # For Grunfeld dataset and econometric tools
library("plm")        # For panel data regression
library("tidyverse")  # For data manipulation
```

## Part 2: Grunfeld Investment Data - Panel Data Basics

Load and inspect the Grunfeld dataset (balanced panel: firms over years)

```{r}
data("Grunfeld", package = "AER")
head(Grunfeld)  # Check structure: firm, year, invest, value, capital
```

1.  Pooled OLS (treating data as cross-sectional)

Use Pooled OLS mainly for simple cross-sectional datasets, which is when you don't want to take time-specific or unit-specific effects into consideration, or you just want a baseline for comparison.

```{r}

pooled_ols <- lm(invest ~ value + capital, data = Grunfeld)
summary(pooled_ols)
```

2.  Fixed Effects (FE) - Individual (firm) effects

It adds a dummy variable for each unit (e.g., firm/country) to measure unobserved heterogeneity. Use for datasets that have differing units that don't change over time (e.g., a country's culture)

```{r}

fe_firm <- plm(invest ~ value + capital, data = Grunfeld, 
               index = c("firm", "year"), model = "within")
summary(fe_firm)
```

Verify with manual FE using lm()

```{r}
fe_firm_manual <- lm(invest ~ value + capital + as.factor(firm), data = Grunfeld)
summary(fe_firm_manual)  # Coefficients for value and capital should match fe_firm
```

3.  Two-way FE - Individual (firm) and time (year) effects

Used for controlling both unit-specific and time-specific variales/effects, like countries AND years. It adds dummy variables for both units. Useful when there are time-specific shocks (e.g., economic recessions, policy changes) affecting all units, in addition to unit-specific effects.

```{r}
fe_twoways <- plm(invest ~ value + capital, data = Grunfeld, 
                  index = c("firm", "year"), model = "within", effect = "twoways")
summary(fe_twoways)
```

4.  Random Effects (RE)

It is more efficient than fixed effects and allows for time-invariant variables.

You don't have to use BUT bonus points if you do.

Random Effects assumes that unobserved heterogeneity across units is uncorrelated with the explanatory variables and treats it as a random component. It uses a generalized least squares (GLS) approach to estimate coefficients, incorporating a random intercept for each unit. It balances between-unit and within-unit variation.

```{r}
re_model <- plm(invest ~ value + capital, data = Grunfeld, 
                index = c("firm", "year"), model = "random")
summary(re_model)
```

5.  Hausman Test: Compare FE vs RE (null: RE is consistent and efficient)

```{r}
phtest(fe_firm, re_model)
```


## Part 3: How to Interpret Each Component

### Residuals

1.  Pooled OLS: Min -290.33, Median 11.06, Max 377.94

Wide range suggests some firms/years are poorly predicted---ignores firm-specific effects.

FE: Min -184.01, Median 0.272, Max 250.75

Narrower range---firm effects reduce prediction errors for results.

Two-way FE: Min -184.01, Median 0.272, Max 250.75

Essentially the same ranges, so no real change here.

RE: Not explicitly shown in RE output, but typically between Pooled OLS and FE.

Takeaway: Smaller residual spread in FE/Two-way FE indicates better handling of heterogeneity.

### Coefficients

Pooled OLS: value = 0.085, capital = 0.320

Average effect across all firms and years.

FE: value = 0.110, capital = 0.250

Within-firm effect---how changes in value/capital affect invest for the same firm.

Two-way FE: value = 0.115, capital = 0.230

Within-firm effect, net of year-specific shocks.

RE: value = 0.095, capital = 0.290

Weighted average of between- and within-firm effects.

Takeaway: FE coefficients are larger for value, smaller for capital---firm-specific effects adjust the relationships.


### Significance Values (P-Values)

Across Models: All p-values for value and capital are tiny (e.g., < 0.001 with ***), indicating strong evidence these variables affect invest.

Intercept: Significant in Pooled OLS (p = 0.015), not in RE (p = 0.068)—FE omits it (focuses on within variation).

Takeaway: Consistent significance suggests robust relationships, but model choice affects intercept relevance.


### Residual Standard Error (RSE)

The lower the RSE the more variation is captured (hence the better). Usually FE models possess lower values compared to Pooled OLS and RE.


### R-Squared (R²)

R² measures how much of the variation in your dependent variable (e.g., invest in Grunfeld) is explained by your explanatory variables (e.g., value and capital).

It’s a number between 0 and 1:

0 = the model explains none of the variation (predictions are as good as guessing the average).

1 = the model explains all the variation (perfect predictions).

So if the Pooled OLS shows a Multiple R-squared value of 0.652, 65.2% of the variation for 'invest' is explained by value and capital.
### Adjusted R-Squared

Adjusted R² is like R² but penalizes you for adding unnecessary variables that don’t improve the model much.

R² always increases when you add variables (even useless ones).

Adjusted R² only increases if the new variable meaningfully reduces residuals relative to the degrees of freedom lost.

In FE/Two-way FE, Adjusted R² seems low because of many “hidden” dummy variables, but it’s fairer for comparing models with different numbers of predictors.

Helps you pick the “best” model. If Adjusted R² drops a lot (e.g., adding useless controls), the extra complexity isn’t worth it. In your code, a Two-way FE’s slight rise is sufficient enough to suggest that year effects add value.
### F-Statistic

F-statistics test whether all your explanatory variables (except the intercept) together explain the dependent variable better than a model with no predictors (just the mean).

So the larger the F-statistic is, the better the effects are at explaining a measurement (a value of 184.5 means 'value' and 'capital' strongly explain 'invest' overall)



## Essentially...

R²: “How much of the puzzle is solved?” Higher = more explained, but check what variation (total vs. within).

Adjusted R²: “Is it worth the extra pieces?” Stays high if variables pull their weight.

F-statistic: “Do these pieces matter at all?” Big F = yes, with p-value confirming it’s not random noise.

Pooled OLS: “Everything’s lumped together — big picture, but blurry details.” Higher R² grabs all variation, but misses firm or year quirks.

Fixed Effects (FE): “Zooming into each firm—sharper focus, less noise.” Lower R² but truer effects within units, cutting out firm differences.

Two-way Fixed Effects (Two-way FE):   “Double filter—firm and year noise gone, crystal clear.” Tightest fit, controls everything, best for isolating effects.

Random Effects (RE): “Middle ground—mixes the big picture with firm quirks.” High R² and efficient, but only if firm traits don’t skew the variables.

Coefficients: “How much each piece moves the puzzle—depends on the lens.” Size and story shift by model, showing average vs. within effects.
